Predicting the conformational preferences of proteins using a physics-based free energy method

### 2013-03-06 Paper/paper.tex
Started changes to text

\subsection{The confinement method correctly identifies the most native-like predictions from a
subset of CASP predictions}

This method is particularly well suited to pick out the native/native like structure from misfolded or decoys. We have
tested this using different models from CASP (critical assessment of structure
prediction). CASP is a blind test in which different groups world wide apply their
methods to predict the 3D structure of proteins given their sequence. This is done with a strict 3
week limit on each target (3 day for servers) and each group is allowed to submit 5 possible
structures (ranked from best to worst). In our role as assessors during the CASP8 and CASP9 \cite{MacCallum2011}
competition we observed no correlation at all between the ordering of structures submited by the
groups and the real ranking compared to an experimental model \cite{Kryshtafovych2011}. The
consequences of this goes beyond
those five structures; the deeper meaning is that groups producing ensembles of structures are
generating structures that are better than the ones they submit, but they do not know about it. In
fact, when querrying different groups after the results were known, most groups agree that this is
the case. Beyond the CASP problem, this reflects on the modelers ability to correctly rank order
models in many different environments, such as structures for drug design leads or designing more
stable proteins or peptide mimetics such as peptoids. One of the main culprits of this lack of
accuracy is the fact that ranking is often done via a potential energy function, which in many cases
lacks an entropy component. Other initiatives including knowledge based potentials do use some sort
of free energy to rank order, but its accuracy is not enough. Our method provides a physics based
solution to this problem. We have tried two experiments centered on CASP. First we tried to rank
order some structures from the previous CASP9 experiment. We have tested different options for that purpose.
First, we rank order submitions from same group and then from different groups. Our second test was 
to check whether this method can do better quality assessment/metaprediction than the best group in that catagory in CASP9. 
In both methods, our ranking is determined as a free energy measure,
and it is compared with the ranking given by a geometrical comparison between native and submitted
models called GDT\_TS (Global distance test score \cite{Zemla2003}). 
GDT\_TS represents the average percentage of residues that are in close proximity in two
structures optimally superimposed using four different distance cutoffs (1, 2, 4 and 8 \AA).

We have chosen the inital models on which to test the methods based on server groups that have
traditionally done good in past CASP events. There is a high correlation between the structures that
are chosen for this method and its predictive accuracy. In particular, when GDT\_TS scores are below
50, it is difficult for us to say anything about the models. Our approach is simple, for every
subset of structures we choose, we use the confinement method in a pairwise procedure, and then rank
the structures. In all cases, we have also calculated the free energy with respect to the native
structure to check the stability of the models.
However, this step is not needed if we just want to rank order.

\subsubsection{The confinement method can rank different models generated by the same methodology and distinguish the native structure}

@@

%\subsubsection{Difficulty in CASP Experiments}

%It is often difficult to model CASP targets using the confinement and release method. There are many cases,
%where the sequence of a particular target is trimmed during actual result presentation. This is illustrated
%with the crystal structure of the leucine zipper domain of cGMP dependent protein kinase I (pdb id: 3NMD
%and CASP9 code: T0605) as shown in Figure~\ref{fig:T0605}. The actual CASP 9 target consisted of 72 amino acid residues, but
%as only 18-66 amino acid residues could be detected in actual crystal structure of this protein, the final result was trimmed 
%with respect to that crystal structure with 49 amino acid residues which consist of a single alpha helix. The amino acid residues 18 and 66 are shown in
%three models submitted by the group "Baker". It is important to note that blind prediction of model ranking on the basis of free energy can be hugely 
%affected in such scenario as the only major difference between those models are the trimmed parts, specially the region consisting residues 1-18.

\subsubsection{Can the confinement method perform better quality assessment in protein structure prediction?}
A part of the CASP experiment is dedicated to the assessment of the quality of predicted models \cite{Kryshtafovych2011}. 
Most of the top performing groups in this category use consensus approaches for quality assessment \cite{Wang2011}.
A review of CASP9 quality assessment also pointed out that the methods which are based on clustering
techniques perform
much better compared to those methods which are based on analysis of individual models \cite{Kryshtafovych2011}. 
In this background we try to answer, whether the confinement method can do better quality assessment
compared to the best performing
group (MUFOLD-WQA) in this catagory. In CASP9, the overall reliability of models in quality assessment (QA) was accepted in a mode known as
QMODE 1. Here, predictors were asked to score each model on a scale from 0 to 1, with higher values corresponding to better models \cite{Kryshtafovych2011}. 
We have again picked up the CASP taget T0538 to answer this question. We have chosen Model 3 submitted by PconsR (GDT\_TS = 96, QMODE 1 = 0.5434) 
and model 5 from the MULTICOM-NOVEL ( GDT\_TS = 83, QMODE 1 = 0.5865). Both are server predicted models and they were further used for quality assessment 
by the group MUFOLD-WQA \cite{Wang2011}. It is important to note that the model from PconsR was the most accurate predicted model for this 
target and the model from MULTICOM-NOVEL was predicted as best by the quality assessment experiments of MUFOLD-WQA. Also, 
the QMODE 1 value presented here is from the group MUFOLD-WQA. 
Our free energy analysis predict that the model from Pconsr is indeed better than the model from 
MULTICOM-NOVEL $(\Delta \Delta G (Model\_Pconsr - Model\_MULTICOM-NOVEL = -3.9 Kcal/Mol)$, which   
matches well with the GDT\_TS score. On the other hand the MUFOLD-WQA predicted the ranking in the reverse order.

@@

\Alberto{The last 50 years represent our effort to understand what the native state of a protein
    looks like by using theoretical methods. Experimental evidence coming from NMR and x-ray
    cristallography remain so far the only real ways to solve the structure of the native state.
    However, this methods are expensive and slow. Theoretical methods on the other hand are cheaper,
    but despite improvement still have a long way to go. Often this computational pipelines produce
    ensembles of diverse structures and identifying which structure is more native like is still a
    problem. Here we have shown how the application of the confinement method can help identify
    native like states. The high success rate balances with the computational expense. However, with
    GPU technology and the rate at which computers improve, this computational expense is becoming
less of a bottle neck.}
\Alberto{I don't agree with some of the points here, but I think I covered what you meant in the
previous paragraph}
In the recent past there have been a number of studies directed towards understanding protein 
structure using theoretical tools. Experimentally, it is relatively easy to identify the native structure from the data obtained 
from x-ray crystallography or NMR. On the other hand using theoretical tools it is often difficult to identify the native structure 
using both physics based methods and knowledge based method. In order to rank structure from the available data, some groups use 
clustering algorithms and the structure of the
most populated cluster is picked out as the most probable structure. However, it may happen that the structure may be trapped and 
spend most of its time in a kinetic trap ***kinetic traps in MD,not other methods, local
minima***, which may give a wrong interpretation of the clustering result. On the other hand, some groups
also depend on potential energy functions to pick out their best structure, lacking an entropic
component. In this work we attempted to identify the native/native like protein strcuture from the decoy using free energy as a scale.
We have used the previously developed confinement method, applying it to larger systems.

We found that in majority of cases, the native structure is always the most stable and we can rank
protein structures upto 100 amino 
acid residues long. This method works very well if at least one of the strcuture in the decoy set
has high quality. The main advantage of this method is 
that it does not require any reaction coordinate. Thus one can calculate the free energy difference between two very different conformations. 
An interesting application from this method that we have exploited is the partition of free energies
on a per residue basis.
 Although, it is an approximate partition, it gives important insight regarding the conformational 
preferences of a particular residue. The use of this decomposition in understanding the effect of a
small number of aminoacids in our test chamaleon sequences allows us to think that in the future,
this method may be used for protein design. 
The method can be also give wrong result if all the models in the decoy are very poor. It may also be difficult to rank structures if 
all the structures are very very close to the native structure. But we want to ignore this fact as our main aim is to differentiate
between the good and the bad structure. In this article we have extensively tested this method using examples from CASP 
This method can be computationally expensive in actual CASP experiment if one want rank hundreds of submitted models in CASP 
experiments. However, in real life experiment with a particular target, we expect this method can give accurate result. The important 
fact about confinement method is that all the part of the calculation can be done independently. 
There is another big bottlenecks in this calculation. This has to do with the amount of
simulations that have to be performed to confine the ensemble of conformations close to a microstate
into a single microstate. We have tackled this issue by using GPU (Graphical Processing Unit) technology as opposed to the classical
CPUs, which gives us a two order of magnitude boost in computational efficiency.
For example, we have mostly carried out our calculation with Amber program \cite{Case2005} in GPU computer. With a avarage 56 residue 
we found that it take only 4 hour to complete 20 ns of the confinement step. So, with available computer power this method will 
be easy to use for real problems. 
%There is another issue, which is normal mode
%analysis calculations are Oh(N**3) and require large amounts of memory, effectively limiting the
%size of the systems under study. In most cases quasiharmonic analysis.

@@

%There are two big bottlenecks in this calculation. The first has to do with the amount of
%simulations that have to be performed to confine the ensemble of conformations close to a microstate
%into a single microstate, the second bottleneck has to do with the size of the system in the normal
%mode calculation. We have tackled the first issue by using GPU (Graphical Processing Unit) technology as opposed to the classical
%CPUs, which gives us a two order of magnitude boost in computational efficiency. Normal mode
%analysis calculations are Oh(N**3) and require large amounts of memory, effectively limiting the
%size of the systems under study. Where as this can be overcome by larger memory machines, these are
%often not available, and so we propose an alternative by using quasiharmonic analysis. This is still
%an Oh(N**3) problem, but we now the motions of atoms are already determined in trajectories, and so
%we can choose to do quasiharmonic analysis excluding hydrogens, which will reduce the number of
%atoms N to almost N/2. This means that we can use this method for systems twice as large.

@@

%\begin{figure}
%\begin{center}
%\includegraphics[width=12cm,height=10cm]{T0605.pdf}
%\end{center}
%\caption{ Three models of the leucine zipper domain of cGMP dependent protein
%kinase I (original pdb id: 3NMD and CASP target: T0605). The original target sequence of this protein was 72 residues.
%However only residues 3 to 66 is only kept during result announcement as that part can be found from crystallographic analysis. The
%region 3 to 66 are almost identical in all the 3 models  }
%\label{fig:T0605}
%\end{figure}

>>>>

\subsection{We have applied the confinement method to structure prediction}

We have applied the confinement method as a ``meta-predictor'' for structure prediction. Here, the
task is to correctly identify the most correct models out of a set of ``decoys'' generated by
different methods during the Critical Assessment of Structure Prediction (CASP) experiment. CASP is
a blind test in which different groups apply methods to predict the 3-dimensional structure of
proteins from their sequences. This is done with a either a three week (for humans) or three day
(for fully automated servers) time limit on each target. Each group is allowed to submit five
possible structures, which they are supposed to rank from best to worst. We have performed two
experiments centered on CASP.

In the first experiment, we examined several targets and tried to rank-order the predictions
produced by other groups. In some cases the structures all came from the same group, in others the
structures were drawn from several different groups. The goal of this experiment is to determine if
the physics-based confinement method can correctly identify more native-like structures as having
low free energies.

Our second experiment was to see if the confinement method can identify structures that are missed
by other meta-prediction servers. Most successful meta-prediction servers are based on the idea of
consensus: if many different prediction methods produce similar results, then that is probably a
correct prediction. This is often a powerful heuristic, but it can miss cases where there is a very
good result that is only predicted by one method. We chose several cases where such structures were
missed and assessed if the confinement method can correctly identify these accurate models.

As is common in the CASP experiment, we assess our results in terms of Global Distance Test Total
Score (GDT-TS) \cite{Zemla2003}, which is a C$\alpha$ based measure of structural accuracy. It can
be understood roughly as the percentage of residues that are correctly positioned in the model
(higher is better, range 0 to 100).

We have chosen the initial models on which to test the methods based on server groups that have
traditionally done well in past CASP events.

\subsubsection{The confinement method can correctly rank different models generated by the same methodology
and distinguish the native structure}

@@

\subsubsection{Can the confinement method perform better quality assessment in protein structure prediction?}

A part of the CASP experiment is dedicated to the assessment of the quality of predicted models
\cite{Kryshtafovych2011}. Most of the top performing groups in this category use consensus
approaches for quality assessment \cite{Wang2011}. A review of CASP9 quality assessment also pointed
out that the methods which are based on clustering techniques perform much better compared to those
methods which are based on analysis of individual models \cite{Kryshtafovych2011}. In this
background we try to answer, whether the confinement method can do better quality assessment
compared to the best performing group (MUFOLD-WQA) in this catagory. In CASP9, the overall
reliability of models in quality assessment (QA) was accepted in a mode known as QMODE 1. Here,
predictors were asked to score each model on a scale from 0 to 1, with higher values corresponding
to better models \cite{Kryshtafovych2011}. We have again picked up the CASP taget T0538 to answer
this question. We have chosen Model 3 submitted by PconsR (GDT\_TS = 96, QMODE 1 = 0.5434) and model
5 from the MULTICOM-NOVEL ( GDT\_TS = 83, QMODE 1 = 0.5865). Both are server predicted models and
they were further used for quality assessment by the group MUFOLD-WQA \cite{Wang2011}. It is
important to note that the model from PconsR was the most accurate predicted model for this target
and the model from MULTICOM-NOVEL was predicted as best by the quality assessment experiments of
MUFOLD-WQA. Also, the QMODE 1 value presented here is from the group MUFOLD-WQA. Our free energy
analysis predict that the model from Pconsr is indeed better than the model from MULTICOM-NOVEL
$(\Delta \Delta G (Model\_Pconsr - Model\_MULTICOM-NOVEL = -3.9 Kcal/Mol)$, which   matches well
with the GDT\_TS score. On the other hand the MUFOLD-WQA predicted the ranking in the reverse order.

@@

\Alberto{The last 50 years represent our effort to understand what the native state of a protein
    looks like by using theoretical methods. Experimental evidence coming from NMR and x-ray
    cristallography remain so far the only real ways to solve the structure of the native state.
    However, this methods are expensive and slow. Theoretical methods on the other hand are cheaper,
    but despite improvement still have a long way to go. Often this computational pipelines produce
    ensembles of diverse structures and identifying which structure is more native like is still a
    problem. Here we have shown how the application of the confinement method can help identify
    native like states. The high success rate balances with the computational expense. However, with
    GPU technology and the rate at which computers improve, this computational expense is becoming
less of a bottle neck.}
\Alberto{I don't agree with some of the points here, but I think I covered what you meant in the
previous paragraph}

In the recent past there have been a number of studies directed towards understanding protein
structure using theoretical tools. Experimentally, it is relatively easy to identify the native
structure from the data obtained from x-ray crystallography or NMR. On the other hand using
theoretical tools it is often difficult to identify the native structure using both physics based
methods and knowledge based method. In order to rank structure from the available data, some groups
use clustering algorithms and the structure of the most populated cluster is picked out as the most
probable structure. However, it may happen that the structure may be trapped and spend most of its
time in a kinetic trap ***kinetic traps in MD,not other methods, local minima***, which may give a
wrong interpretation of the clustering result. On the other hand, some groups also depend on
potential energy functions to pick out their best structure, lacking an entropic component. In this
work we attempted to identify the native/native like protein strcuture from the decoy using free
energy as a scale. We have used the previously developed confinement method, applying it to larger
systems.

We found that in majority of cases, the native structure is always the most stable and we can rank
protein structures upto 100 amino acid residues long. This method works very well if at least one of
the strcuture in the decoy set has high quality. The main advantage of this method is that it does
not require any reaction coordinate. Thus one can calculate the free energy difference between two
very different conformations. An interesting application from this method that we have exploited is
the partition of free energies on a per residue basis. Although, it is an approximate partition, it
gives important insight regarding the conformational preferences of a particular residue. The use of
this decomposition in understanding the effect of a small number of aminoacids in our test chamaleon
sequences allows us to think that in the future, this method may be used for protein design. The
method can be also give wrong result if all the models in the decoy are very poor. It may also be
difficult to rank structures if all the structures are very very close to the native structure. But
we want to ignore this fact as our main aim is to differentiate between the good and the bad
structure. In this article we have extensively tested this method using examples from CASP This
method can be computationally expensive in actual CASP experiment if one want rank hundreds of
submitted models in CASP experiments. However, in real life experiment with a particular target, we
expect this method can give accurate result. The important fact about confinement method is that all
the part of the calculation can be done independently. There is another big bottlenecks in this
calculation. This has to do with the amount of simulations that have to be performed to confine the
ensemble of conformations close to a microstate into a single microstate. We have tackled this issue
by using GPU (Graphical Processing Unit) technology as opposed to the classical CPUs, which gives us
a two order of magnitude boost in computational efficiency. For example, we have mostly carried out
our calculation with Amber program \cite{Case2005} in GPU computer. With a avarage 56 residue we
found that it take only 4 hour to complete 20 ns of the confinement step. So, with available
computer power this method will be easy to use for real problems. 
%There is another issue, which is normal mode analysis calculations are Oh(N**3) and require large
%amounts of memory, effectively limiting the size of the systems under study. In most cases
%quasiharmonic analysis.

@@

%There are two big bottlenecks in this calculation. The first has to do with the amount of
%simulations that have to be performed to confine the ensemble of conformations close to a
%microstate into a single microstate, the second bottleneck has to do with the size of the system in
%the normal mode calculation. We have tackled the first issue by using GPU (Graphical Processing
%Unit) technology as opposed to the classical CPUs, which gives us a two order of magnitude boost in
%computational efficiency. Normal mode analysis calculations are Oh(N**3) and require large amounts
%of memory, effectively limiting the size of the systems under study. Where as this can be overcome
%by larger memory machines, these are often not available, and so we propose an alternative by using
%quasiharmonic analysis. This is still an Oh(N**3) problem, but we now the motions of atoms are
%already determined in trajectories, and so we can choose to do quasiharmonic analysis excluding
%hydrogens, which will reduce the number of atoms N to almost N/2. This means that we can use this
%method for systems twice as large.

@@
